---
title: "PHP 1510 Midterm Exam 2019"
date: "Due: October 29, 2019 at 11:59pm "
output: pdf_document
---



<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
options(digits=3)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
```


\vspace{1cm}

Name: _______________________________________

\vspace{2cm}


# Instructions

- Write your code and answers under each of the following question. 
- You may use course notes, labs and homeworks for this exam. 
- You may not ask any person for any help. 
- If you are found copying code or responses from anyone, you will receive a 0 for the entire exam. 
- **Explain your reasoning**. The final answer is not as important as the process.
- All interpretations must be in context to the original problem including units. 
- **All answers must be in complete sentences**
- Download the Rmd document to run with your code such. 
- Please submit to: ***[PHP 1510: Canvas](https://canvas.brown.edu/courses/1078851/assignments/7746089)***


\pagebreak

# Scoring

------------------------------------------------
Problem       Point Value         Problem Grade
------------  ------------------  --------------
                                  `
                                  
1             3                  `____________`


                                  `

2             3                  `____________`

                                  `

3             4                  `____________`

                                  `

4             4                  `____________`

                                  `

5             6                  `____________`

                                  `
                                  
6             6                  `____________`

                                  `

7             6                  `____________`

                                  `
                                  
8             6                  `____________`

                                  `
                                  
9             12                  `____________`

                                  `
                                  
10            20                 `____________`

                                  `
                                  
                                  `                                  `
              
              **Total**           70
----------------------------------------------------------

\pagebreak



# The Data

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in
the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC).
The BRFSS is administered and supported by CDC's Population Health Surveillance Branch, under the Division
of Population Health at the National Center for Chronic Disease Prevention and Health Promotion. The BRFSS
is an ongoing surveillance system designed to measure behavioral risk factors for the noninstitutionalized adult
population (aged 18 years of age and older) residing in the United States. The BRFSS was initiated in 1984,
with 15 states collecting surveillance data on risk behaviors through monthly telephone interviews. Over time,
the number of states participating in the survey increased, and by 2001, 50 states, the District of Columbia,
Puerto Rico, Guam, and the US Virgin Islands were participating in the BRFSS. Today, all 50 states, the
District of Columbia, Puerto Rico, and Guam collect data annually; American Samoa, the Federated States of
Micronesia, and Palau collect survey data over a limited point-in-time (usually 1 to 3 months). In this document,
the term state is used to refer to all areas participating in the BRFSS, including the District of Columbia, Guam,
and the Commonwealth of Puerto Rico. 

Factors assessed by the BRFSS in 2014 include tobacco use, HIV/AIDS knowledge and
prevention, exercise, immunization, health status, healthy days health-related quality of life, health care access,
inadequate sleep, chronic health conditions, alcohol consumption, oral health, falls, drinking and driving, cancer
screenings (including breast, cervical, prostate, colorectal cancers), and seatbelt use. Since 2011, the BRFSS
conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline
telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the
cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates
by using a cellular telephone and resides in a private residence or college housing. 


\newpage

----------------------------------------
Variable    Description 
--------------- ----------------------------
genhlth         Excellent
            
                Very Good
            
                Good
            
                Fair
            
                Poor

genhlth_bin     Excellent/Very Good/Good

                Fair/Poor
            
Unhealthy.days  0-30

menthlth        0-30

poorhlth        0-30

imprace         White, Non-Hispanic

                Black, Non-Hispanic
                
                Asian, Non-Hispanic
                
                AIAN, Non-Hispanic
                
                Hispanic
                
                Other Race, Non-Hispanic

insurance       yes

                No
                
trnsgndr        Yes, mtf

                Yes, ftm
                
                Yes, non-conforming
                
                No

trnsgndr_bin    Yes

                No
                
sxorient        Heterosexual

                Homosexual
                
                Bisexual
                
                Other
                
sxorient_bin    Heterosexual

                Other
                
lstisfy         Very Satisfied

                Satisfied
                
                Disatisfied
                
                Very Disatisfied
                
lsatisfy_bin    Satisfied

                Disatisfied
                
emtsuprt        Always

                Usually
                
                Sometimes
                
                Rarely
                
                Never
                
emtsuprt_bin    Always/Usually

                Sometimes/Rarely/Never
                
medcost         Yes

                No
--------------------------------------------------------------

-----------------------------------------

\pagebreak


## Conceptual Questions


1. **(3 points)** What does the expecation tell you?

\vspace{4.5cm} 





2. **(3 points)** What does the variance tell you?


\vspace{5cm}


3. **(4 points)** What is the story behind Bernoulli Data?

\vspace{5cm}


4. **(4 points)**  What are some traits of the normal distribution?


\newpage





## Data Questions


Our goal here will to be consider the days that stress, depression, and problems with emotions caused days where mental health was not good. We will be addressing this specifically in context to the transgender experience. The reason for this being that we do see different rates of depression, suicide and other outcomes associated with the transgender experience. It is important to begin to understand what the cause might be. 



- Download the file ***[`brfss.rda`](https://drive.google.com/file/d/188YvZMXQxegZY5oXDoeTa8Lkh8hGbzx4/view?usp=sharing)***

***Click the link or go to: https://drive.google.com/file/d/188YvZMXQxegZY5oXDoeTa8Lkh8hGbzx4/view?usp=sharing***

```{r, echo=FALSE}

load("brfss.rda")
```

5. **(6 points)** Plot and Describe the Distribution of poor mental health days. 

```{r}
## Place code for #5 here
library(tidyverse)
library(ggplot2)
```


\vspace{5cm}



6. **(6 points)** Plot and Describe the Distribution of Race. 


```{r}
## Place code for #6 here

```


\vspace{5cm}


7. **(6 points)** Plot and Describe the Distribution of Transgender Binary. 

```{r}
## Place code for #7 here

```


\vspace{5cm}

8. **(6 points)** Plot and Describe the Distribution of General Health (not binary).


```{r}
## Place code for #8 here

```

\newpage

9. **(12 points)** Display graphs of variables that have relationships with poor mental health days.  (**Note:** Look at the notes with general health to see how to combine multiple plots in a larger grid image. (lect 13 slide 58) )

```{r}
## Place code for #9 here
library(gridExtra)

```

\newpage

10. **20 points** We have been exploring the outcome of days off due to mental health. Papers have been published suggesting that there is a relationship between the transgender experience and mental health outcomes. 

Health related to the transgender experience in general has not been well researched. This is an attempt to get everyone to be thinking about the number of days in which stress, depression and other emotions may had led to poor mental health days.

Do there appear to be relationships between the transgender experience and the number of days in which stress, depression and emotions led to poor mental health? 

***Remember: we have 3 things to consider with a hypothesis.***

***Note: This problem is much more complex than we have data to consider, this is one starting place with data which is available to us.***

***Note: You will need to use transgender binary to make this work like the other examples.***



