---
title: "PHP 2510 Midterm Exam 2019"
date: "Due: October 29, 2019 at 11:59pm "
output: pdf_document
---



<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
options(digits=3)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
```


\vspace{1cm}

Name: _______________________________________

\vspace{2cm}


# Instructions

- Write your code and answers under each of the following question. 
- You may use course notes, labs and homeworks for this exam. 
- You may not ask any person for any help. 
- If you are found copying code or responses from anyone, you will receive a 0 for the entire exam. 
- **Explain your reasoning**. The final answer is not as important as the process.
- All interpretations must be in context to the original problem including units. 
- **All answers must be in complete sentences**
- Download the Rmd document to run with your code such. 
- Please submit to: ***[PHP 2510: Canvas](https://canvas.brown.edu/courses/1078852/assignments/7746091)***


\pagebreak

# Scoring

------------------------------------------------
Problem       Point Value         Problem Grade
------------  ------------------  --------------
                                  `
                                  
1             3                  `____________`


                                  `

2             3                  `____________`

                                  `

3             4                  `____________`

                                  `

4             4                  `____________`

                                  `

5             5                  `____________`

                                  `
                                  
6             6                  `____________`

                                  `

7             6                  `____________`

                                  `
                                  
8             6                  `____________`

                                  `
                                  
9             6                  `____________`

                                  `
                                  
10            12                 `____________`

                                  `
                                  
11            20                 `____________`

                                  `                                  
                                  `                                  `
              
              **Total**           75
----------------------------------------------------------

\pagebreak



# The Data

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in
the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC).
The BRFSS is administered and supported by CDC's Population Health Surveillance Branch, under the Division
of Population Health at the National Center for Chronic Disease Prevention and Health Promotion. The BRFSS
is an ongoing surveillance system designed to measure behavioral risk factors for the noninstitutionalized adult
population (aged 18 years of age and older) residing in the United States. The BRFSS was initiated in 1984,
with 15 states collecting surveillance data on risk behaviors through monthly telephone interviews. Over time,
the number of states participating in the survey increased, and by 2001, 50 states, the District of Columbia,
Puerto Rico, Guam, and the US Virgin Islands were participating in the BRFSS. Today, all 50 states, the
District of Columbia, Puerto Rico, and Guam collect data annually; American Samoa, the Federated States of
Micronesia, and Palau collect survey data over a limited point-in-time (usually 1 to 3 months). In this document,
the term state is used to refer to all areas participating in the BRFSS, including the District of Columbia, Guam,
and the Commonwealth of Puerto Rico. 

Factors assessed by the BRFSS in 2014 include tobacco use, HIV/AIDS knowledge and
prevention, exercise, immunization, health status, healthy days health-related quality of life, health care access,
inadequate sleep, chronic health conditions, alcohol consumption, oral health, falls, drinking and driving, cancer
screenings (including breast, cervical, prostate, colorectal cancers), and seatbelt use. Since 2011, the BRFSS
conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline
telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the
cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates
by using a cellular telephone and resides in a private residence or college housing. 


\newpage

----------------------------------------
Variable    Description 
--------------- ----------------------------
genhlth         Excellent
            
                Very Good
            
                Good
            
                Fair
            
                Poor

genhlth_bin     Excellent/Very Good/Good

                Fair/Poor
            
Unhealthy.days  0-30

menthlth        0-30

poorhlth        0-30

imprace         White, Non-Hispanic

                Black, Non-Hispanic
                
                Asian, Non-Hispanic
                
                AIAN, Non-Hispanic
                
                Hispanic
                
                Other Race, Non-Hispanic

insurance       yes

                No
                
trnsgndr        Yes, mtf

                Yes, ftm
                
                Yes, non-conforming
                
                No

trnsgndr_bin    Yes

                No
                
sxorient        Heterosexual

                Homosexual
                
                Bisexual
                
                Other
                
sxorient_bin    Heterosexual

                Other
                
lstisfy         Very Satisfied

                Satisfied
                
                Disatisfied
                
                Very Disatisfied
                
lsatisfy_bin    Satisfied

                Disatisfied
                
emtsuprt        Always

                Usually
                
                Sometimes
                
                Rarely
                
                Never
                
emtsuprt_bin    Always/Usually

                Sometimes/Rarely/Never
                
medcost         Yes

                No
--------------------------------------------------------------

-----------------------------------------

\pagebreak


## Conceptual Questions


1. **(3 points)** What does the expecation tell you?

\vspace{4.5cm} 





2. **(3 points)** What does the variance tell you?


\vspace{5cm}


3. **(4 points)** What is the story behind Bernoulli Data?

\vspace{5cm}


4. **(4 points)**  What are some traits of the normal distribution?


\newpage


5. **(5 points)** Does the story below fit the Binomial Distribution? If not, why does it not fit the story. 

Bass dwell in a particular lake. There are N bass, of which a simple random sample of size n are caught and tagged ("simple random sample" means that all sets of n bass are equally likely). The caught bass are returned to the population, and then a new sample is drawn, this time with size m. This is an important method that is widely-used in
ecology, known as capture-recapture. 

Is the probability of exactly k of the m bass in the new sample were previously tagged binomial? (Assume that a bass that was caught before
doesn't become more or less likely to be caught again.)
\vspace{8cm}


\newpage

## Data Questions

- Download the file ***[`brfss.rda`](https://drive.google.com/file/d/188YvZMXQxegZY5oXDoeTa8Lkh8hGbzx4/view?usp=sharing)***

***Click the link or go to: https://drive.google.com/file/d/188YvZMXQxegZY5oXDoeTa8Lkh8hGbzx4/view?usp=sharing***

```{r, echo=FALSE}

load("brfss.rda")
```

6. **(6 points)** Plot and Describe the Distribution of Mental Health Days off. 

```{r}
## Place code for #6 here
library(tidyverse)
library(ggplot2)
```


\vspace{5cm}



7. **(6 points)** Plot and Describe the Distribution of Race. 


```{r}
## Place code for #7 here

```


\vspace{5cm}


8. **(6 points)** Plot and Describe the Distribution of Transgender Binary. 

```{r}
## Place code for #8 here

```


\vspace{5cm}

9. **(6 points)** Plot and Describe the Distribution of General Health (not binary).


```{r}
## Place code for #9 here

```

\newpage

10. **(12 points)** Display graphs of variables that have relationships with Mental Health Days off.  (**Note:** Look at the notes with general health to see how to combine multiple plots in a larger grid image. (lect 13 slide 58) )

```{r}
## Place code for #10 here
library(gridExtra)

```

\newpage

11. **20 points** We have been exploring the outcome of days off due to mental health. Papers have been published suggesting that there is a relationship between transgender and mental health. Some of this comes from the fact that there are higher transgender suicide attempts. 
Transgender health in general has not been well researched and so one good place to start might be to consider the impact of how many days mental health issues were bad. 
Do you find that those who classify as transgender have higher number of bad mental health days? 

***Remember: we have 3 things to consider with a hypothesis.***

***Note: This problem is much more complex than we have data to consider, this is one starting place with data which is available to us.***

***Note: You will need to use transgender binary to make this work like the other examples.***



