---
title: PHP 1510/2510 Lab 3
subtitle: Sampling Distributions in R
---


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=F)
knitr::opts_chunk$set(eval=F)
```




# Central Limit Theorem with Binary Outcome

- Consider a response where we are interested in a whether or not a population has had it. 
- Let's assume that we have 30% of the population with a "Yes" and that leaves 70% with "No". 
- Then if we were to sample from this population, `Yes=1` and `No=0`. 
- This means the mean of these samples represent the proportion of "Yes" responses. 



For Each of the code below, make sure you interpret what you are seeing below the output. 




## Binomial samples with sample size of 10 each time. 

### 10 Samples of 10 Each

```{r clt-bin-10, echo=F}
# get 10 samples with 10 scores each


t=10
n=10

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p1 <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)

p1
```




### 100 Samples of 10 Each

```{r clt-bin-100, echo=F}
# get 100 samples with 10 scores each


t=100
n=10 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p2 <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)

p2
```


### 1000 Samples of 10 Each

```{r clt-bin-1000, echo=F}
# get 1000 samples with 10 scores each


t=1000
n=10 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p3 <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means")  + xlim(-1,1)

p3
```




### 10000 Samples of 10 Each

```{r clt-bin-10000, echo=F}
# get 10000 samples with 10 scores each


t=10000
n=10 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p4 <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)
p4

```



### Progression

```{r clt-bin-prog-10, echo=F, fig.height=5, fig.width=12}
library(gridExtra)
grid.arrange(p1,p2,p3,p4, ncol=4)
```





### Binomial samples with sample size of 30 each time. 

- We will be looking at the same population with 30% "Yes".
- This time we will do samples of 30 each. 

### 10 Samples of 30 Each

```{r clt-bin-10b, echo=F}
# get 10 samples with 30 scores each


t=10
n=30

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p1b <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)

p1b
```


### 100 Samples of 30 Each

```{r clt-bin-100b, echo=F}
# get 100 samples with 30 scores each


t=100
n=30 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p2b <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)

p2b
```


### 1000 Samples of 30 Each

```{r clt-bin-1000b, echo=F}
# get 1000 samples with 30 scores each


t=1000
n=30 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p3b <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means")  + xlim(-1,1)

p3b
```


### 10000 Samples of 30 Each

```{r clt-bin-10000b, echo=F}
# get 10000 samples with 30 scores each


t=10000
n=30 

scores <- rbinom(t*n, 1, 0.3)
samples <- rep(1:t,each=n)
my_df <- data.frame(samples,scores)

# get the means of the samples

sample_means <- my_df %>%
                group_by(samples) %>%
                summarise(means=mean(scores))

# make a histogram

p4b <-  ggplot(sample_means, aes(x=means))+
  geom_histogram(color="white")+
  theme_classic() + xlab("Means") + xlim(-1,1)
p4b

```

### Progression

```{r clt-bin-prog-10b, echo=F, fig.height=5, fig.width=12}
library(gridExtra)
grid.arrange(p1b,p2b,p3b,p4b, ncol=4)
```


### Progression of Both

```{r clt-bin-prog-10c, echo=F, fig.height=5, fig.width=12}
grid.arrange(p1, p2, p3, p4, p1b, p2b, p3b, p4b, ncol=4, nrow=2)
```




# Confidence Intervals


## Bonmial Data


### Using Normal Distribution

```{r}
data <- rbinom(10, 1, 0.3)

mn <- mean(data)
std.dev <- sqrt(10*0.3*0.7)/10

low = mn - 1.96*std.dev
high = mn + 1.96*std.dev

low
high
```


- When can we use this? 


### Using t-distribution

```{r}

std.dev = sd(data)

t <- qt(0.975, df=length(data)-1)


low = mn - t*std.dev
high = mn + t*std.dev

low
high
```


### Bootstrapping



```{r}
library(rsample)
library(purrr)
library(ggplot2)
library(dplyr)

set.seed(1234)
bt_data <- bootstraps(data, times = 1000)

bt_data

```


- What information are we given on the splits?

```{R}
bt_data$splits[[1]]
```


#### The Function

```{r}
get_mean <- function(split) {
  # access to the sample data 
  split_data <- analysis(split)
  # calculate the sample mean value 
  split_mean <- mean(split_data$menthlth, na.rm=T)
  return(split_mean)
}
```


#### Pass the data through function:

```{R}
bt_data$bt_means <- map_dbl(bt_data$splits, get_mean)
bt_data
```



#### Graph Distribution

```{R}

ggplot(bt_data, aes(x = bt_means)) + 
  geom_line(stat = "density") + 
  xlab("Mean of Poor Mental Health Days")
```


#### Bootstrap Confidence Intervals

```{r}
bt_ci <- round(quantile(bt_data$bt_means, c(0.025, 0.975)), 3)
bt_ci
```



- How do these compare? 



#### Graphical Display of Bootstrap Interval

```{r}
bt_ci_lower <- quantile(bt_data$bt_means, c(0.025))
bt_ci_upper <- quantile(bt_data$bt_means, c(0.975))
bt_mean <- mean(bt_data$bt_means)
 
bt_data <- bt_data %>% 
  dplyr::mutate(Color = ifelse(bt_means < bt_ci_lower | bt_means > bt_ci_upper, "Out of 95% CI", "In 95% CI"))
 
 
ggplot(bt_data, aes(x = bt_means, y = 0)) + 
  geom_jitter(aes(color = Color), alpha = 0.6, size = 3, width = 0) + 
  geom_vline(xintercept = round(c(bt_ci_lower, bt_mean, bt_ci_upper),2), linetype = c(2,1,2)) + 
  scale_x_continuous(breaks = round(c(bt_ci_lower, bt_mean, bt_ci_upper),2),
                     labels = c("Lower CI (95%)", "Mean", "Upper CI (95%)"),
                     sec.axis = sec_axis(~., breaks = round(c(bt_ci_lower, bt_mean, bt_ci_upper),2),
                                         labels = round(c(bt_ci_lower, bt_mean, bt_ci_upper),2))) + 
  scale_color_manual(values = c("gray40", "firebrick4")) + 
  theme_void() + 
  theme(legend.title = element_blank(),
        legend.position = "top", 
        legend.text = element_text(size = 14), 
        axis.text.x = element_text(size = 14))
```


- Modify the code to use with the other 2 styles of confidence intervals. How do these 3 plots compare? 
```{r}
