---
title       : Transforming Data with `dplyr`
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

```{r, echo=F}
require(tidyverse)
```



# The `dplyr` Package


--- .class #id

## The `dplyr` Package

- When we work with data, we need to be able to clean it and modify it as needed.  
- We may wish to  do things like:
    - add additional variables.
    - Scale variables
    - Collapse variables 
- The `dplyr` package allows us to further work with our data. 

--- .class #id


## `dplyr` Functionality

- With `dplyr` we have five basic verbs that we will learn to work with:
  - `filter()`
  - `select()`
  - `arrange()`
  - `mutate()`
  - `summarize()`


  


---  .segue bg:grey


# Filtering 


--- .class #id

## Filtering


- At this point we will consider how we pick the rows of the data that we wish to work with. 
- If you consider many modern data sets, we have so much information that we may not want to bring it all in at once. 
- R brings data into the RAM of your computer. This means you can be limited for what size data you can bring in at once. 
- Very rarely do you need the entire data set. 
- We will focus on how to pick the rows or observations we want now.


--- .class #id

## Enter the `filter()` Function

- The `filter()` function chooses rows that meet a specific criteria. 
- We can do this with Base R functions or with   `dplyr`. 

```{r}
library(dplyr)
```

--- .class #id

## The Data

- We will begin by looking at some data from [FiveThirtyEight](https://fivethirtyeight.com/)
- They have a wealth of data stored with an R package and lots of data analysis done online. 
- We will work with a dataset behind the story ["Comic Books Are Still Made By Men, For Men And About Men"](https://fivethirtyeight.com/features/women-in-comic-books/).
- *I know this is not public health but it has some interesting attributes to the data*.



--- .class #id

## Exploring the Data

- It is always good to learn how to explore the data and we will learn this more as we go. 
- We will start with a simple command to look at the structure of the data. 


--- .class #id

## Structure of the Data

```{R, eval=F}
library(tidyverse)
library(fivethirtyeight)

str(comic_characters)
```


--- .class #id

## Structure of the Data

```{R, echo=F, message=F}
library(fivethirtyeight)

str(comic_characters)
```



--- .class #id

## Filtering Example

- Filtering is the idea of considering only certain observations. 
- THis means filtering is done at the observation/person level.
- Let's say we only want to consider "Marvel Characters"
- Traditionally in R, this was done with what is called indexing but we will use the `dplyr` tools for this. 

--- .class #id

## Filtering Example

```{r,eval=F}
library(tidyverse)
comic_characters %>%
  filter(publisher=="Marvel")
```

--- .class #id

## Filtering Example

```{r,echo=F, message=F}
library(tidyverse)
comic_characters %>%
  filter(publisher=="Marvel")
```

--- .class #id

## More than One Variable Filtered

- We can also filter on more than just one variable. 
- We could filter by characters before 1970 and from Marvel. 

```{r, eval=F}
comic_characters %>%
    filter(publisher=="Marvel" & year<1970)
```


--- .class #id

## More than One Variable Filtered

- We can also chain together more than one filter command and achieve the same results:

```{r, eval=F}
comic_characters %>%
    filter(publisher=="Marvel") %>%
    filter(year<1970)
```


--- .class #id

## Results

```{R, echo=F}
comic_characters %>%
    filter(publisher=="Marvel") %>%
    filter(year<1970)
```




--- .class #id


## Further Filtering

- `filter()` supports the use of multiple conditions where we can use Boolean. 
- For example if we wanted to consider only observations that have a  first year of between 1980 and 1985:

```{r, eval=FALSE}
comic_characters %>% 
    filter(year>=1980 & year<=1985)
```

--- .class #id


## Further Filtering

```{r, echo=FALSE}
comic_characters %>% 
    filter(year>=1980 & year<=1985)
```

--- .class #id


## Further Filtering

- We can also use the `filter()` function to remove missing data for us. 
- On Datacamp you learned a class of functions called `is.`*foo*`()` where *foo* represents a data type. 
- We could choose to only use observations that have gender stated. 
- That means we wish to not have missing data for gender. 


```{r, eval=FALSE}
comic_characters %>% filter(!is.na(sex))
```

--- .class #id


## On Your Own: RStudio Practice

- These slides are for you to practice on your own when you are at home. 
- I will state the problem and data on one slide.
- On the next I will show what your answer should look like. 
- Code for the solutions can be found on github. 

--- .class #id

## On Your Own: RStudio Practice

- For these data we will consider the gapminder dataset. 
- You will need the `gapminder` package.
- You can run `install.packages("gapminder")` to get it. 

--- .class #id 

## On Your Own: RStudio Practice

Using the `filter()` function and chaining: 

- Choose only rows associated with
  - Uganda
  - Rwanda

--- .class #id


## On Your Own: RStudio Practice


Your end result should be:

```{R, echo=FALSE}
library(gapminder)
filter(gapminder, country=="Uganda" | country=="Rwanda")
```

---  .segue bg:grey



# Selecting


--- .class #id

## Selecting

- The next logical step would be to select the columns we want as well. 
- Many times we have so many columns that we are not interested in for a particular analysis. 
- Instead of slowing down your analysis by continuing to run through extra data, we could just select the columns we care about. 

--- .class #id


## Enter the `select()` Function

- The `select()` function chooses columns that we specify. 
- Again we can do this with base functions or with `dplyr`. 
- We feel that as you continue on with your R usage that you will most likely want to go the route of `dplyr` functions instead.



--- .class #id

## Selecting Example 

- If we explore the comics data further, we can see that there may be some variables we are more interested than others. 
- Consider character attributes, we may wish to just consider name, gender, eye color and hair color. 

We could then do the following

```{r,eval= FALSE}
comic_characters %>%
    select(name, sex, eye, hair)
```


--- .class #id

## Selecting Example Continued


```{r,echo= FALSE}
comic_characters %>%
    select(name, sex, eye, hair)
```


--- .class #id

## Removing Columns


- We may wish to pick certain columns that we wish to have but we also may want to remove certain columns. 
- It is quite common to de-identify a dataset before actually distributing it to a research team. 
- The `select()` function will also remove columns. 



--- .class #id

## Removing Columns

- Lets say that we wished to remove the `page_id` and `urlslug` of the characters:

```{R, eval=FALSE}
comic_characters %>%
  select(-page_id,-urlslug)
```

--- .class #id

## Removing Columns

We also could use a vector for this:

```{R, eval=F}

cols <- c("url_slug", "page_id")
gapminder %>%
  select(-one_of(cols))
```



--- .class #id

## On Your Own: RStudio Practice

- Consider the gapminder data: `gapminder`. 
  1. Select all but the `pop` column.
  2. Remove the year and lifeExp from them.
  3. Select values which contain "p" in them (hint: look up `contains()` function). 
  4. Chain these together so that you run a command and it does all of these things. 

--- .class #id

## On Your Own: RStudio Practice


Your answer should look like:

```{R, echo=FALSE}
gapminder %>% select(-pop)  %>% select(-lifeExp,-year) %>% select(contains("p"))
```



--- .segue bg:grey

# Arranging the Data

--- .class #id

## Arranging the Data

- We also have need to make sure the data is ordered in a certain manner. This can be easily done in R with the `arrange()` function. 
- Again we can do this in base R but this is not always a clear path. 

--- .class #id

## Arranging the Data Example

- Let's say that we wish to look characters year of entry and total appearances. 
- Then say further we want to sort it by appearances. 


--- .class #id

## Arranging the Data Example Continued

```{r, eval=FALSE}

comic_characters %>%
    select(name,year,  appearances) %>%
    arrange(appearances)
```

--- .class #id

## Arranging the Data Example Continued

```{r, echo=FALSE}

comic_characters %>%
    select(name,year,  appearances) %>%
    arrange(appearances)
```

--- .class #id

## Arranging the Data Example Continued

- With `arrange()` we first use `select()` to pick the only columns that we want and then we arrange by the `appearances`. 
- If we had wished to order them in a descending manner we could have simply used the `desc()` function:

```{r, eval=FALSE }
comic_characters %>%
    select(name,year,  appearances) %>%
    arrange(desc(appearances))
```

--- .class #id


## More Complex Arrange

- Lets consider that we wish to look at the top 3 Appearances for each year.
- Then we wish to order them from largest to smallest Appearances. 
- We then need to do the following:
    1. Group by year.
    2. Pick the top 3 appearance.
    3. order them largest to smallest.
    



--- .class #id

## More Complex Arrange Continued


```{R, eval=FALSE}
comic_characters %>% 
  group_by(year) %>%  
  top_n(3, appearances) %>% 
  arrange(desc(appearances))
```

- Where
    - `group_by()` is a way to group data. This way we perform operations on a group. So top 3 appearances are grouped by year. 
    - `top_n()`takes a tibble and returns a specific number of rows based on a chosen value. 


--- .class #id

## More Complex Arrange Continued


```{R, echo=FALSE}
comic_characters %>% 
  group_by(year) %>%  
  top_n(3, appearances) %>% 
  arrange(desc(appearances))
```



--- .class #id

## On Your Own: RStudio Practice

- Perform the following operations:
  - Group by year. 
  - use `sample_n()` to pick 1 observation per year 
  - Arrange by longest to smallest life expectancy. 



--- .class #id





## On Your Own: RStudio Practice


Your answer ***may*** look like:

```{r, echo=FALSE}
gapminder %>%
  group_by(year) %>%
  sample_n(1) %>%
  arrange(desc(lifeExp))
```

---  .segue bg:grey


# Adding New Variables


--- .class #id

## Adding New Variables

- There is usually no way around needing a new variable in your data. 
- For example, most medical studies have height and weight in them, however many times what a researcher is interested in using is Body Mass Index (BMI). 
- We would need to add BMI in. 


--- .class #id

## Adding New Variables

- Using the `tidyverse` we can add new variables in multiple ways
  - `mutate()`
  - `transmute()`


--- .class #id

## Example

- Consider appearances, this can be deceiving as some characters have been around for 50 years more than others. 
- A more logical way might be to consider average appearances per year. 
- This means we could create 2 variables for this. 
    - Total Years
    - Average Appearances per year. 

--- .class #id

## Adding Appearances


- We can do this with the `mutate()` function. 

```{r, eval=FALSE}
comic_characters %>%
    select(name, year, appearances) %>%
    mutate(total_years = 2014-year) %>%
    mutate(avg_appearances = appearances/total_years) %>%
    arrange(desc(total_years))
```

--- .class #id

## Example: Mutate

```{r, echo=FALSE}
comic_characters %>%
    select(name, year, appearances) %>%
    mutate(total_years = 2014-year) %>%
    mutate(avg_appearances = appearances/total_years) %>%
    arrange(desc(total_years))
```


--- .class #id

## Changing the sort

```{r, echo=FALSE}
comic_characters %>%
    select(name, year, appearances) %>%
    mutate(total_years = 2014-year) %>%
    mutate(avg_appearances = appearances/total_years) %>%
    arrange(total_years)
```


--- .class #id

## What can we see? 

- Our initial premise of more appearances is based on more time may not be accurate when weighting by year.
- Why?
    - Some might be characters that died and we didnt calculate this. 
    - Some may have been introduced and then dropped. 
    - Maybe more comics are produced now than in 1935. 