---
title       : Summarization of Data
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

```{r, echo=F}
require(tidyverse)
```


# Summarizing Data

--- .class #id


## Piping or Chaining

- We will discuss a concept that will help us greatly when it comes to working with our data. 
- The usual way to perform multiple operations in one line is by nesting. 

--- .class #id

## Piping or Chaining

To consider an example we will look at the data provided in the gapminder package:

```{r}
library(gapminder)
head(gapminder)
```

--- .class #id

## Nesting vs Chaining

- Let's say that we want to have the GDP per capita and life expectancy Kenya.
- Traditionally speaking we could do this in a nested manner:


```{r, eval=F}
filter(select(gapminder, country, lifeExp, gdpPercap), country=="Kenya")
```

--- .class #id

## Nesting vs Chaining

- It is not easy to see exactly what this code was doing but we can write this in a manner that follows our logic much better. 
- The code below represents how to do this with chaining. 

```{r, eval=F}
gapminder %>%
    select(country, lifeExp, gdpPercap) %>%
    filter(country=="Kenya")
```

--- .class #id

## Breaking Down the Code

- We now have something that is much clearer to read.
- Here is what our chaining command says:

1. Take the `gapminder` data
2. Select the variables: `country`, `lifeExp` and `gdpPercap`.
3. Only keep information from Kenya. 

- The nested code says the same thing but it is hard to see what is going on if you have not been coding for very long.


--- .class #id

## Breaking Down the Code

- The result of this search is below: 


```
# A tibble: 12 x 3
   country lifeExp gdpPercap
    <fctr>   <dbl>     <dbl>
 1   Kenya  42.270  853.5409
 2   Kenya  44.686  944.4383
 3   Kenya  47.949  896.9664
 4   Kenya  50.654 1056.7365
 5   Kenya  53.559 1222.3600
 6   Kenya  56.155 1267.6132
 7   Kenya  58.766 1348.2258
 8   Kenya  59.339 1361.9369
 9   Kenya  59.285 1341.9217
10   Kenya  54.407 1360.4850
11   Kenya  50.992 1287.5147
12   Kenya  54.110 1463.2493
```

--- .class #id

## What is `%>%`

- In the previous code we saw that we used `%>%` in the command you can think of this as saying ***then***.
- For example:

```{r, eval=F}
gapminder %>%
    select(country, lifeExp, gdpPercap) %>%
    filter(country=="Kenya")
```

--- .class #id

## What Does this Mean?

- This translates to:
    - Take Gapminder ***then*** select these columns select(country, lifeExp, gdpPercap) ***then*** filter out so we only keep Kenya


--- .class #id

## Why Chain?

- We still might ask why we would want to do this. 
- Chaining increases readability significantly when there are many commands. 
- With many packages we can replace the need to perform nested arguments. 
- The chaining operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package. 

--- .class #id

## User Defined Function


- Let's say that we wish to find the Euclidean distance between two vectors say, `x1` and `x2`. 
- We could use the math formula:

\[\sqrt{sum(x1-x2)^2}\]

--- .class #id

## User Defined Function

- In the nested manner this would be:

```{r results='hide'}
x1 <- 1:5; x2 <- 2:6
sqrt(sum((x1-x2)^2))
```

--- .class #id

## User Defined Function


- However, if we chain this we can see how we would perform this mathematically. 
```{r, eval=FALSE}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```
- If we did it by hand we would perform elementwise subtraction of `x2` from `x1` ***then*** we would sum those elementwise values ***then*** we would take the square root of the sum. 


--- .class #id

## User Defined Function


```{r}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```

- Many of us have been performing calculations by this type of method for years, so that chaining really is more natural for us. 



--- .class #id

## Summarizing Data

- As you have seen in your own work, being able to summarize information is crucial. 
- We need to be able to take out data and summarize it as well. 
- We will consider doing this using the `summarise()` function. 

--- .class #id

## Summarizing Data

- We will begin with a simple summary:
    1. Create a table grouped by `gender`.
    2. Summarize each group by taking mean of `appearances`.

--- .class #id

## Renaming Variable

- First we need to correct the data and rename the variable `sex` to `gender`. 

```{r}
library(fivethirtyeight)
comic_characters <- comic_characters %>%
                      mutate(gender=sex)
```


--- .class #id

## Summarizing Data Example 


```{r, eval=FALSE}

comic_characters %>%
    group_by(gender) %>%
    summarise(avg_appear = mean(appearances, na.rm=TRUE))
```



--- .class #id

## Summarizing Data Example 


```{r, echo=FALSE}
comic_characters %>%
    group_by(gender) %>%
    summarise(avg_appear = mean(appearances, na.rm=TRUE))
```

--- .class #id


## Another Example

- Lets say that we would like to have more than just the averages but we wish to have the minimum and the maximum year of entry by gender:

```{r, eval=FALSE}

comic_characters %>%
    group_by(gender) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), year)
```


--- .class #id


## Another Example



```{r, echo=FALSE}
comic_characters %>%
    group_by(gender) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), year)
```

--- .class #id

## On Your Own: RStudio Practice 

- The following is a new function:
  - Helper function `n()` counts the number of rows in a group
- Then for each year and continent:
  - count total lifeExp
  - Sort in descending order. 

--- .class #id

## On Your Own: RStudio Practice 

Your answer should look like:

```{r, echo=FALSE}

gapminder %>%
    group_by(continent,year) %>%
    summarise(lifeExp_count = n()) %>%
    arrange(desc(lifeExp_count))
```


We could also have used what is called the  `tally()` function:

```{r}
gapminder %>%
    group_by(country, year) %>%
    tally(sort = TRUE)
```

--- .class #id

## Further Summaries

- We have so far discussed how one could find the basic number summaries:
  - mean
  - median
  - standard deviation
  - variance
  - minimum 
  - maximum
- However there are many more operations that you may wish to do for summarizing data. 
- In fact many of the following examples are excellent choices for working with categorical data which does not always make sense to do the above summaries for. 


--- .class #id

## Further Summaries

- We will consider:
  1. Grouping and Counting
  2. Grouping, Counting and Sorting
  3. Other Groupings
  4. Counting Groups


--- .class #id



## Grouping and Counting


- We have seen the functions `tally()` and `count()`. 
- Both of these can be used for grouping and counting. 
- They also are very concise in how they are called. 


--- .class #id



## Grouping and Counting


- For example if we wished to know how many characters there were by year, we would use `tally()` in this manner: 

```{r, eval=FALSE }
comic_characters %>%
  group_by(year) %>% 
  tally()
```

--- .class #id



## Grouping and Counting

- Where as we could do the same thing with `count()`

```{r, eval=FALSE}
comic_characters %>% 
  count(year)
```

*Notice: `count()` allowed for month to be called inside of it, removing the need for the `group_by()` function. 


--- .class #id



## Grouping, counting and sorting.

- Both `tally()` and `count()` have an argument called `sort()`. 
- This allows you to go one step further and group by, count and sort at the same time. 
- For `tally()` this would be:

```{r, eval=FALSE}
comic_characters %>% 
    group_by(year) %>% 
    tally(sort=TRUE)
```

--- .class #id

## Grouping, counting and sorting: `tally()`

```{r, echo=FALSE}
comic_characters %>% 
    group_by(year) %>% 
    tally(sort=TRUE)
```




--- .class #id

## Grouping with other functions

- We can also sum over other values rather than just counting the rows like the above examples. 
- For example let us say we were interested in knowing the total appearances for characters in a given gender and publisher.
- We could do this with the `summarise()` function, `tally()` function or the `count()` function:

```{r, eval=FALSE}
comic_characters %>% 
  group_by(gender, publisher) %>% 
  summarise(total_app = sum(appearances))
```


--- .class #id

## Grouping with other functions


```{r, echo=FALSE}
comic_characters %>% 
  group_by(gender, publisher) %>% 
  summarise(total_app = sum(appearances))
```


--- .class #id



--- .class #id

## Grouping with other functions

- For  `tally()` we could do:

```{r ,eval=FALSE}
comic_characters %>% 
  group_by(gender, publisher) %>% 
  tally(wt = appearances)
```

*Note: in `tally()` the `wt` stands for weight and allows you to weight the sum based on the appearances*. 

--- .class #id

## Grouping with other functions

- With the `count()` function we also use `wt`:


```{r}
comic_characters %>% count(gender, publisher, wt = appearances)
```

--- .class #id

## Counting Groups


- We may want to know how large our groups are. To do this we can use the following functions:
  - `group_size()` is a function that returns counts of group. 
  - `n_groups()` returns the number of groups

--- .class #id

## Counting Groups

- So if wanted to count the number of characters by hair color, we could group by hair color and find the groups size using `group_size()`:

```{r, eval=FALSE}
comic_characters %>% 
  group_by(hair) %>% 
  group_size()
```


--- .class #id

## Counting Groups



```{r, echo=FALSE}
comic_characters %>% 
  group_by(hair) %>% 
  group_size()
```


--- .class #id

## Counting Groups


- If we just wished to know how many years were represented in our data we could use the `n_groups()` function:


```{r, eval=FALSE}
comic_characters %>% 
  group_by(year) %>% 
  n_groups()
```



--- .class #id

## Counting Groups

```{r, echo=FALSE}
comic_characters %>% 
  group_by(year) %>% 
  n_groups()
```

