---
title       : Categorical Data
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=FALSE)
library(ggplot2)
library(fivethirtyeight)
require(tidyverse)
library(broom)
library(dplyr)
```

# Categorical Data Analysis

--- .class #id

## Categorical Data Analysis

- So far we have focused primarily on continuous outcomes and how to compare them. 
- Many times we have categorical data that we wish to compare. 
- We have seen this with a proportion test but there are other methods to doing this. 

--- .class #id

## Chi-Square Distribution

- This is a crucial distribution for working with categorical data and many other things. 
- This distribution is specifically from the standard normal distribution. 

--- .class #id

## Chi-Square Distribution

- Consider $V$, where

$$V=Z_1^2 + \cdots + Z_n^2$$

- Where
    - $Z_i$ is a standard normal distribution. 
    - Each $Z_i$ are independent from each other. 

Then we say that $V \sim \chi^2_n$


--- .class #id

## Chi Square Distribution

```{R, echo=F}
x <- seq(0.1,15, by =0.001)
plot(x, dchisq(x,1), type="l", col="black", lwd=3, lty=2)
curve(dchisq(x, 2), add=TRUE, col="red", lwd=3)
curve(dchisq(x, 3), add=TRUE, col="blue", lwd=3)
curve(dchisq(x, 5), add=TRUE, col="green", lwd=3)
curve(dchisq(x, 10), add=TRUE, col="purple", lwd=3)
legend(10,1.2,legend=c("df=1","df=2", "df=3", "df=5", "df=10"),lty=1,col=c("black", "red", "blue", "green", "purple"))
```

--- .class #id

## Some Data

- We have some data from a study done in 1894-96 on Cholera Innoculations in Calcutta, India. 
- Of the 818 exposed subjects, 279 receive an innoculation and 539 did not. 

|     | Infected | Not Infected |  | 
| -------- | --------- | ---------- | ----------- | 
| Innoculated | 3 | 276 | 279  | 
| Not Innoculated | 66 | 473 | 539 | 
|  | 69 | 749 | 818 | 


--- .class #id

## What can we do? 

- We could look at proportions:

$$Pr(\text{Probability of Cholera}| \text{Innoculated}) = \dfrac{3}{279}= 0.011$$
$$Pr(\text{Probability of Cholera}| \text{Not Innoculated}) = \dfrac{66}{539}= 0.122$$

- How do we decide? 

--- .class #id

## Testing Proportions

1.Test $H_0: p_1=p_2 \text{ vs } H_1: p_1\ne p_2$
2. Test $H_0: OR=1 \text{ vs } H_1: OR\ne 1$
3. Pearson $\chi^2$ Test


--- .class #id

## Pearson Chi-square Test

$$X = \sum_{i,j} \dfrac{\left(O_{ij} - E_{ij} \right)^2}{E_{ij}}$$

- Where
    - $O_{ij}$: observed count in cell $i,j$
    - $E_{ij}$: expected count in cell $i,j$
    - $x \sim \chi^2(df=(r-1)(c-1))$

--- .class #id

## Our Data


```{r}
M <- as.table(rbind(c(3, 276), c(66, 473)))
dimnames(M) <- list(Treatment = c("Innoculated", "Not Innoculated"),
                    Status = c("Infected", "Not Infected"))
M
```

--- .class #id

## Chi Square Test

```{r}
chisq.test(M, correct=FALSE)
```

--- .class #id

## Attrition Data

```{R}
attr_pop <- rsample::attrition
```

--- .class #id

## Gender vs Attrition

```{r}
table(attr_pop$Gender, attr_pop$Attrition)
```

--- .class #id

## Chi Square Test

```{r}
T <- table(attr_pop$Gender, attr_pop$Attrition)
chisq.test(T, correct=F)
```

--- .class #id

## Gender vs Performance Rating

```{r gender-perfor, echo=F}
library(ggplot2)
ggplot(attr_pop, aes(x=PerformanceRating, fill=Gender)) + 
 geom_bar()
```
 
--- .class #id

## Gender vs Performance Rating 
 
```{r}
table(attr_pop$Gender, attr_pop$PerformanceRating)
```


--- .class #id

## Gender vs Performance Rating

```{r}
T <- table(attr_pop$Gender, attr_pop$PerformanceRating)
chisq.test(T, correct=F)
```



--- .class #id

## Gender vs Performance Rating

```{r}
attr_pop2 <- attr_pop %>%
    filter(PerformanceRating=="Excellent" | PerformanceRating == "Outstanding")
attr_pop2 <- droplevels(attr_pop2)
T <- table(attr_pop2$Gender, attr_pop2$PerformanceRating)
chisq.test(T, correct=F)
```

--- .class #id

## Work Life Balance and Attrition

```{r worklife, echo=F}
ggplot(attr_pop, aes(x=Attrition, fill=WorkLifeBalance)) + geom_bar()
```

--- .class #id

## Work Life Balance and Attrition

```{r}
table(attr_pop$WorkLifeBalance, attr_pop$Attrition)
```


--- .class #id

## Work Life Balance and Attrition

```{r}
T <- table(attr_pop$WorkLifeBalance, attr_pop$Attrition)
chisq.test(T, correct=F)
```

--- .class #id

## Work Life Balance and Attrition

```{r, eval=F}
table(attr_pop$WorkLifeBalance, attr_pop$Attrition) %>%
prop.table() %>%
round(., 2)
```

--- .class #id

## Work Life Balance and Attrition

```{r, echo=F}
table(attr_pop$WorkLifeBalance, attr_pop$Attrition) %>%
prop.table() %>%
round(., 2)
```



--- .class #id

## Work Life Balance and Attrition

```{r}
attr_pop %>%
    count(WorkLifeBalance, Attrition) %>%
    mutate(freq = n/sum(n)) %>%
    select(-n) %>%
    spread(WorkLifeBalance, freq)
```




--- .class #id

## Work Life Balance and Attrition

```{r}
attr_pop %>%
    count(WorkLifeBalance, Attrition) %>%
    mutate(freq = n/sum(n)) %>%
    select(-n) %>%
    spread(Attrition, freq)
```



--- .class #id

## Work Life Balance and Attrition

```{r}
attr_pop %>%
    count( Attrition, WorkLifeBalance) %>%
    mutate(freq = n/sum(n)) %>%
    select(-n) %>%
    spread(Attrition, freq)
```



---  .segue bg:grey

# ANOVA

--- .class #id


## Recap on our progress:
  
- Up to this point we have seen basic statistical features:
  - mean
  - variance
  - standard deviation
  - median
  - min 
  - max
- We have also considered a t-test in which we compare a continuous variable across 2 groups. 

--- .class #id 

## Enter ANOVA

- What if we need to compare more than 2 groups? 
- Lets say we have the groups: A, B and C
- We could compare:
  - A vs B
  - A vs C
  - B vs C
- What are some issues with this?

--- .class #id 

## Enter ANOVA

- Multiple Testing issues
    - Each time you perform this test on the same data, you use a type 1 error of 0.05, the more tests you perfrom the more this error increases. 
    - If you need to do multiple testing you then have to do a p-value correction. 
- Could be a waste of time if all the groups are the same. 
    - More computational time if there is no difference. 


--- .class #id

## Enter ANOVA

- We then can consider ANOVA: ANalysis Of VAriance
- ANOVA asks a very basic question:
    - Where is the variability coming from?
        - Is it coming from within each group?
        - Is it coming between the groups? 
- The hypothesis test we perform is 

$$ H_0:\;\; \mu_1 = \mu_2 = \cdots = \mu_k$$ 
$$ H_1:\;\; \text{At least one group is different}$$



--- .class #id

## What is our test then? 

- We now move to testing with the $F$ distribution rather than the $t$-distribution:
- We use the following test statistic:
$$ k = \dfrac{\text{Measure of Between-Group Variability}}{\text{Measure of Within-Group Variability}}$$ 
- How do we calculate these variabilities?


--- .class #id

## The math

- We calculate the following values:
    - Between Sum of Squares
$$SS_B = \sum_{i=1}^k n_i\left(\bar{y}_i - \bar{y}\right)^2$$
    - Within Sum of Squares
$$SS_W = \sum_{i=1}^k \sum_{j=1}^{n_i} \left(y_{ij} - \bar{y}_i\right)^2$$
    - Total Sums of Squares
$$SS = SS_B + SS_W$$
- In all of these: $i$ is the index for $k$ groups and $j$ is the index for the $n_i$ observations in each group. 



--- .class #id

## ANOVA Variances

- This is where the analysis of variances comes in, we are comparing the variances:
    - Between group variability
    - Within group variability
- Traditionally this test was performed using the following table:


--- .class #id 

## ANOVA Table

|                     |   DF  |    Sum Sq.   |       Mean Sq       |           F value           |  Pr(>F) |
|:-------------------:|:-----:|:------------:|:-------------------:|:---------------------------:|:-------:|
| Between (treatment) | $k-1$ |  $SS_{B}$  |      $MS_{B} = \dfrac{SS_B}{k-1}$     | $\frac{MS_{B}}{MS_{W}}$ | p-value |
|    Within (error)   | $N-k$ | $SS_{W}$   | $MS_{W} = \dfrac{SS_W}{N-k}$ |                             |         |
|    Total  | $N-1$ | $SS_{T}$   |  |                             |         |


--- .class #id

## Calculating ANOVA

- In a traditional class you would be made to do this by hand
- We won't do this:
    1. I hate doing it. 
    2. you never do this in real life. 
    3. WE HAVE R!
- We will focus on performing this in R. 


--- .class #id

## The Data for Class

- We will consider the data behind the story: ["Comic Books are Still Made By Men, For Men and About Men"](http://fivethirtyeight.com/features/women-in-comic-books/).
- This data is part of the `fivethirtyeight` package:
- To explore the variable names run the following code:

```{r, eval=F}
library(fivethirtyeight)
?comic_characters
```


--- .class #id


## Difference in Appearances by Gender

- Lets consider if the number of appearances of characters is different depending on the gender of the character. 
- We could first graph this:

```{r, eval=F}
library(ggplot2)
ggplot(comic_characters, aes(x = sex, y = appearances)) +
  geom_point()  + 
  geom_point(stat = "summary", fun.y = "mean", color = "red", size = 3)
```



--- .class #id

## Difference in Appearances by Gender

```{r, echo=F}
library(ggplot2)
ggplot(comic_characters, aes(x = sex, y = appearances)) +
  geom_point()  + 
  geom_point(stat = "summary", fun.y = "mean", color = "red", size = 3)
```

--- .class #id

## Table of Gender

- We can see that this is hard to read, we can see what the groups look like by counting them 

```{r, eval=F}
library(tidyverse)
comic_characters %>%
    group_by(sex) %>%
    tally(sort = TRUE)
```


--- .class #id

## Table of Gender

- We can see that this is hard to read, we can see what the groups look like by counting them 

```{r, echo=F}
library(tidyverse)
comic_characters %>%
    group_by(sex) %>%
    tally(sort = TRUE)
```

--- .class #id 

## Data Cleaning

- We can make the names smaller

```{r}
comic <- comic_characters %>%
      mutate(sex = fct_recode(sex, 
      "Agender" = "Agender Characters",
      "Female" = "Female Characters", 
      "Genderfluid" = "Genderfluid Characters",
      "Genderless" = "Genderless Characters",
      "Male" = "Male Characters",
      "Transgender" = "Transgender Characters"
      ))
```

--- .class #id

## Data Cleaning


```{r, echo=F}
ggplot(comic, aes(x = sex, y = appearances)) +
  geom_point()  + 
  geom_point(stat = "summary", fun.y = "mean", color = "red", size = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


--- .class #id 

## Cleaning Data

- we can see that we do not have many people in categories asside from "Male" and "Female"
- This can be a problem with many statistical tests so we can combine categories

```{r}
comic <- comic_characters %>%
      mutate(sex = fct_recode(sex, 
      "Non-Binary" = "Agender Characters",
      "Female" = "Female Characters", 
      "Non-Binary" = "Genderfluid Characters",
      "Non-Binary" = "Genderless Characters",
      "Male" = "Male Characters",
      "Non-Binary" = "Transgender Characters"
      ))
```


--- .class #id 

## Cleaning Data

- We can also see that we have a lot of values that seem to be very high compared to the mean. 
- In this case many times we pull in the extreme values with a log transform
- We can do this with mutate

```{r}
comic <- comic %>%
    mutate(log_app = log(appearances))
```

--- .class #id

## Boxplots

- we can then try looking at boxplots

```{r, echo=F}
ggplot(comic, aes(x = sex, y = log_app)) +
  geom_boxplot()  + 
  geom_point(stat = "summary", fun.y = "mean", color = "red", size = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


--- .class #id 

## Finally ANOVA

- these look a little better now that we have done a log transform
- We can code an ANOVA in r with the following:
  
```{r}
aov(log_app~sex, data=comic)
```


--- .class #id

## What can we do to get more information

- Many things in R are stored in objects called lists. 
- Lists contain a large amount of objects

```{r}
my_anova <- aov(log_app~sex, data=comic)
names(my_anova)
```

--- .class #id 

## Summary

- The summary function works with anova and many other functions to give us a basic summary

```{r}
summary(my_anova)
```



--- .class #id

## What were we testing again?

- Recall our hypothesis:
$$ H_0:\;\; \mu_1 = \mu_2 = \cdots = \mu_k$$ 
$$ H_1:\;\; \text{At least one group is different}$$
- What can we say about these groups?


--- .class #id

## What is Next?

- Now that we know there is a difference, we need to find out what difference that is. 
- This does leave us with a multiple testing problem. 
- Previously it was mentioned that performing multiple hypothesis tests we have problems with the type 1 error. 
- Type 1 error is the error of making a mistake by rejecting the null hypothesis when you shouldn't have. 
- This means that if we perform 20 studies we can assume that we made a mistake on 5% of them or 1 of them will be significant and lead to rejecting the null hypothesis. 


--- .class #id

## What about Multiple Testing

- When we perform 20 tests on the same data what we have is:
$$\begin{align*}
\Pr(\text{At least 1 Significant Result}) &= 1 - \Pr(\text{No Significant Results})\\
&= 1 - \left( 1-0.05\right)^{20}\\
& = 0.6415141
\end{align*}$$
- We call this the **Family Wise Error Rate (FWER)**
- So now we have around 13 tests that we would be making a mistake on. 
- On Friday, this will be covered more thoroughly. 


--- .class #id

## What Type of Multiple Tests for ANOVA

- We need to control the FWER so that $\text{FWER}\le0.05$. 
- There are various methods out there:
    - Bonferroni Method
    - Tukey HSD
    - Holm, Hommel, Dunnett, Šidák , ...


--- .class #id

## The Bonferroni  Correction

- Consider the problem of testing $n$ different tests. - We can do the Bonferroni in 2 different ways:
    - Adjust the significance level
$$ \alpha^*  = \dfrac{\alpha}{n}$$
    - Bonferroni Correct p-values
$$\text{min}\left[2\times\binom{k}{2}\times \Pr\left(\mid t\mid < t_{n-k}\right),1\right]$$


--- .class #id

## The Bonferroni  Correction

- If we have $n=20$ then i we wish to control the FWER at $\alpha=0.05$, then we have
$$\alpha^* = \dfrac{\alpha}{n}= \dfrac{0.05}{20}=0.0025$$
- What does this mean for the FWER:
$$\begin{align*}
\Pr(\text{At least 1 Significant Result}) &= 1 - \Pr(\text{No Significant Results})\\
&= 1 - \left( 1-0.0025\right)^{20}\\
& = 0.04883012
\end{align*}$$


--- .class #id

## Bonferonni in R

- We can perform multiple t-tests in R using:
```
pairwise.t.test(x,g,p.adjust.method,...)
```
- Where
    - `x` is the response vector
    - `g` is the grouping factor
    - `p.adjust.method` is p-value adjustment
    - `...` Others you can see in r


--- .class #id

## Bonferonni in R

- Perform multiple tests
```{R}
attach(comic)
pairwise.t.test(log_app,sex, p.adjust="none")
detach()
```
- Remember to compare vs $\alpha=0.05/3\approx0.017$


--- .class #id

## Bonferonni in R

- Or correct for Bonferroni in the p-values
```{R}
attach(comic)
pairwise.t.test(log_app,sex, p.adjust="bonferroni")
detach()
```


--- .class #id

## Tukey HSD Test

- This is called the Tuker Honest Significant Difference (HSD) test.
- This creates a set of confidence intervals and adjust p-values based on the *studentized range distribution*. 
- Tukey's is usually preferred in ANOVA as it is less conservative that Bonferroni and in many cases yields and exact correction. 

--- .class #id

## Tukey HSD in R

```{r}
TukeyHSD(my_anova, conf.level=0.95)
```

--- .class #id

## Results

- What can we confirm from these tests?


--- .class #id

## Assumptions of ANOVA

- There are assumptions made for every statistical method.
- The assumptions of ANOVA are:
    - Independent groups
    - Homogeneity of Variances
    - Normality of residuals
