---
title       :  Multiple Linear Regression
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(results="hold")
knitr::opts_chunk$set(cache=FALSE)
library(ggplot2)
library(fivethirtyeight)
require(tidyverse)
library(broom)
```



# Multiple Linear Regression 



--- .class #id




## Multiple Regression

- We have been discussing simple models so far. 
- This works well when you have:
    - Randomized Data to test between specific groups (Treatment vs Control)
- In most situations we need look at more than just one relationship. 
- Think of this as needing more information to tell the entire story. 

--- .class #id



## Multiple Linear Regression with appearances



- First start with univariate models
- Then perform the multiple model





--- .class #id

## Multivariate Models


```{r}
mod3 <- lm(appearances~publisher + year, data=comic_characters)
tidy3 <- tidy(mod3, conf.int=T)[,-c(3:4)]
tidy3
```

--- .class #id

## Interpreting Multiple Coefficients

- The intercept is when all coefficients are zero. 
- Each other coefficient is interpreted in context to another. 


--- .class #id

## Interpreting Multiple Coefficients: Our Example

- Intercept: DC average appearances at year 0. ***Not Meaningful***
- Publisher Coefficient: If we consider 2 characters in the same year, the character from Marvel will have on average 9.54 less appearances than the character from DC. 
- Year Coefficient: If we consider 2 characters from the same publisher, an increase in 1 year will lead to on average 0.62 less appearances. 



--- .segue bg:grey


# Multiple Regression


--- .class #id

## Multiple Regression

- We have been discussing simple models so far. 
- This works well when you have:
    - Randomized Data to test between specific groups (Treatment vs Control)
- In most situations we need look at more than just one relationship. 
- Think of this as needing more information to tell the entire story. 


--- .class #id


## Motivating Example

- Health disparities are very real and exist across individuals and populations. 
- Before developing methods of remedying these disparities we need to be able to identify where there are disparities.In this homework we will consider a study by [(Asch & Armstrong, 2007)](http://www.ncbi.nlm.nih.gov/pubmed/17513818).  
- This paper considers 222 patients with localized prostate cancer. 


--- .class #id



## Motivating Example 

- The table below partitions patients by race, hospital and whether or not the patient received a prostatectomy. 

|       |   Race | Prostatectomy | No Prostatectomy | 
| -------------- | -------- | ---------- | ------------ |
University Hospital | White | 54 | 37 | 
|  | Black | 7 | 5  |
| VA Hospital | White | 11 | 29 | 
| | Black | 22 | 57 | 



--- .class #id






## Loading the Data


You can load this data into R with the code below:


```{r}
phil_disp <- read.table("https://drive.google.com/uc?export=download&id=0B8CsRLdwqzbzOXlIRl9VcjNJRFU", header=TRUE, sep=",")
```


--- .class #id



## The Data 

This dataset contains the following variables: 


| Variable |       Description | 
| ----------- | -------------------- | 
| hospital  |     0 - University Hospital |
|           |     1 - VA Hospital | 
| race   |      0 - White |
|      |        1 - Black | 
| surgery |       0 - No prostatectomy |
|          |    1 - Had Prostatectomy | 
| number    |    Count of people in Category |





--- .class #id

## Consider Prostatectomy by Race

```{r}
library(broom)
prost_race <- glm(surgery ~ race, weight=number, data= phil_disp,
                  family="binomial")
tidy(prost_race, exponentiate=T, conf.int=T)[,-c(3:4)]
```



--- .class #id



## Consider Prostatectomy by Race

- What can we conclude? 
- What kind of policy might we want to invoke based on this discovery?



--- .class #id


## Consider Prostatectomy by Hospital

```{r}
prost_hosp <- glm(surgery ~ hospital, weight=number, data= phil_disp,
                  family="binomial")
tidy(prost_hosp, exponentiate =T, conf.int=T)[,-c(3:4)]
```


--- .class #id


## Consider Prostatectomy by Hospital

- What can we conclude? 


--- .class #id

## Multiple Regression of Prostatectomy


```{r}
prost <- glm(surgery ~ hospital + race, weight=number, data= phil_disp,
             family="binomial")
tidy(prost, exponentiate=T, conf.int=T)[,-c(3:4)]
```




--- .class #id


## Multiple Regression of Prostatectomy

- What can We conclude?
- What happened here?
- Does this change our policy suggestion from before?




--- .class #id


## Benefits of Multiple Regression


- Multiple Regression helps us tell a more complete story. 
- Multiple regression controls for confounding. 




--- .class #id


## Confounding

- Associated with both the Exposure and the Outcome
- Even if the Exposure and Outcome are not related, unmeasured confounding can show that they are. 



--- .class #id



## What Do We Do with Confounding?

- We must add all confounders into our model. 
- Without adjusting for confounders are results may be highly biased. 
- Without adjusting for confounding we may make incorrect policies that do not fix the problem. 



--- .class #id


## Multiple Linear Regression with appearances



- First start with univariate models
- Then perform the multiple model






--- .class #id


## Multivariate Models


```{r}
library(broom)
library(fivethirtyeight)
mod3 <- lm(appearances~publisher + year, data=comic_characters)
tidy3 <- tidy(mod3, conf.int=T)[,-c(3:4)]
tidy3
```



--- .class #id

## Interpreting Multiple Coefficients

- The intercept is when all coefficients are zero. 
- Each other coefficient is interpreted in context to another. 




--- .class #id


## Interpreting Multiple Coefficients: Our Example

- Intercept: DC average appearances at year 0. ***Not Meaningful***
- Publisher Coefficient: If we consider 2 characters in the same year, the character from Marvel will have on average 9.54 less appearances than the character from DC. 
- Year Coefficient: If we consider 2 characters from the same publisher, an increase in 1 year will lead to on average 0.62 less appearances. 



--- .class #id

## Further Example: Bike Sharing Data


- We have hourly data spanning 2 years
- This dataset has the first 19 days of each month. 
- Goal is to find the total count of bike rented. 


--- .class #id


## Further Example: Bike Sharing Data

| Data |  Fields |
| :------: | :----------------------|
| datetime | hourly date + timestamp  |
| season |   1 = spring, 2 = summer, 3 = fall, 4 = winter  |
| holiday |  whether the day is considered a holiday |
| workingday |  whether the day is neither a weekend nor holiday |


--- .class #id


## Further Example: Bike Sharing Data

| Data |  Fields |
| :------: | :----------------------|
| weather |  1: Clear, Few clouds, Partly cloudy, Partly cloudy  |
| | 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist | 
| | 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds |
| | 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog |
| temp |  temperature in Celsius |


--- .class #id


## Further Example: Bike Sharing Data

| Data |  Fields |
| :------: | :----------------------|
| atemp |  "feels like" temperature in Celsius |
| humidity |  relative humidity |
| windspeed |  wind speed|
| casual |  number of non-registered user rentals initiated |
| registered |  number of registered user rentals initiated |
| count |  number of total rentals |


--- .class #id

## Univariate Regressions

```{r}
library(readr)
library(tidyverse)
bikes <- read_csv("../Notes/Data/bike_sharing.csv") %>%
        mutate(season = as.factor(season)) %>%
        mutate(weather=as.factor(weather)) 
       
```


--- .class #id

## Univariate Regressions

```{r}
mod1 <- lm(count~season, data=bikes)
mod2 <- lm(count~holiday, data=bikes)
mod3 <- lm(count~workingday, data=bikes)
mod4 <- lm(count~weather, data=bikes)
mod5 <- lm(count~temp, data=bikes)
mod6 <- lm(count~atemp, data=bikes)
mod7 <- lm(count~humidity, data=bikes)
mod8 <- lm(count~windspeed, data=bikes)
mod9 <- lm(count~casual, data=bikes)
mod10 <- lm(count~registered, data=bikes)
```


--- .class #id

## Univariate Regressions

```{r, eval=F}
library(broom)
tidy1 <- tidy( mod1, conf.int=T )[-1, -c(3:4) ]
tidy2 <- tidy(mod2, conf.int=T )[-1, -c(3:4) ]
tidy3 <- tidy(mod3 , conf.int=T)[-1, -c(3:4) ]
tidy4 <- tidy(mod4 , conf.int=T)[-1, -c(3:4) ]
tidy5 <- tidy(mod5, conf.int=T)[-1, -c(3:4) ]
tidy6 <- tidy(mod6 , conf.int=T)[-1, -c(3:4) ]
tidy7 <- tidy(mod7 , conf.int=T)[-1, -c(3:4) ]
tidy8 <- tidy(mod8 , conf.int=T)[-1, -c(3:4) ]
tidy9 <- tidy(mod9, conf.int=T)[-1, -c(3:4) ]
tidy10 <- tidy(mod10, conf.int=T)[-1, -c(3:4) ]
bind_rows(tidy1, tidy2, tidy3, tidy4, tidy5, tidy6, tidy7, tidy8, tidy9, tidy10) 

```



--- .class #id

## Univariate Regressions

```{r, echo=F}
library(broom)
tidy1 <- tidy( mod1, conf.int=T )[-1, -c(3:4) ]
tidy2 <- tidy(mod2, conf.int=T )[-1, -c(3:4) ]
tidy3 <- tidy(mod3 , conf.int=T)[-1, -c(3:4) ]
tidy4 <- tidy(mod4 , conf.int=T)[-1, -c(3:4) ]
tidy5 <- tidy(mod5, conf.int=T)[-1, -c(3:4) ]
tidy6 <- tidy(mod6 , conf.int=T)[-1, -c(3:4) ]
tidy7 <- tidy(mod7 , conf.int=T)[-1, -c(3:4) ]
tidy8 <- tidy(mod8 , conf.int=T)[-1, -c(3:4) ]
tidy9 <- tidy(mod9, conf.int=T)[-1, -c(3:4) ]
tidy10 <- tidy(mod10, conf.int=T)[-1, -c(3:4) ]
rbind(tidy1, tidy2, tidy3, tidy4, tidy5, tidy6, tidy7, tidy8, tidy9, tidy10)
```


--- .class #id


## Multivariate

```{r, eval=F}
mod.final <- lm(count~season+weather+humidity+windspeed, data=bikes)
tidy(mod.final)[-1,-c(3:4)]
glance(mod.final)
```

--- .class #id

## Multivariate

```{r, echo=F}
mod.final <- lm(count~season+weather+humidity+windspeed, data=bikes)
tidy(mod.final)[-1,-c(3:4)]
```

--- .class #id


## Multivariate

```{r, echo=F}
glance(mod.final)
```

--- .class #id

## Multiple Variables


```{r}
library(readr)
library(tidyverse)
bikes <- read_csv("../Notes/Data/bike_sharing.csv") %>%
        mutate(season = as.factor(season)) %>%
        mutate(weather=as.factor(weather)) 
       
```


--- .class #id

## Multiple Variables

```{r, eval=F}
mod.final <- lm(count~season+weather+humidity+windspeed, data=bikes)
tidy(mod.final)[-1,-c(3:4)]
glance(mod.final)
```

--- .class #id


## Multiple Variables

```{r, echo=F}
mod.final <- lm(count~season+weather+humidity+windspeed, data=bikes)
kable(tidy(mod.final)[-1,-c(3:4)])
```