---
title       : Case Study in Data Exploration and Hypothesis Testing
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

```{r, echo=F}
require(tidyverse)
require(gapminder)
```

# Case Study in Data Exploration and Hypothesis Testing

--- .class #id

## The Data

- We will begin working with data from the Behavioral Risk Factor Surveillance System 2014 data. 
- We have a small segment of data that contains information on health and access to healthcare. 

--- .class #id

## The data

```{r, eval=F}
load("Data/brfss.rda")
str(brfss2)
```


--- .class #id

```{r, echo=FALSE}
load("Data/brfss.rda")
str(brfss2)
```

---  .segue bg:grey

# Codebook

--- .class #id

## General Health

- `genhlth` : Would you say your general health is?
    - Excellent
    - Very Good
    - Good
    - Fair
    - Poor

--- .class #id

## Plot

```{r genhealth, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(genhlth, fill=genhlth)) + geom_bar()
```

--- .class #id

## General Health Binary

- `genhlth_bin` : Would you say your general health is?
    - Excellent/Very Good/Good
    - Fair/Poor

--- .class #id

## Plot

```{r genhealth-binary, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(genhlth_bin, fill=genhlth_bin)) + geom_bar()
```

--- .class #id

## Unhealthy Days

- `unhealthy.days` : Now thinking about your physical health, which includes physical illness and injury, for how many days during the past
30 days was your physical health not good?
- Values: 0-30

--- .class #id

## Plot

```{r unhealthy-days, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(unhealthy.days)) + geom_histogram()
```



--- .class #id

## Mental Health Days

- `menthlth` : : Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many
days during the past 30 days was your mental health not good?
- Values: 0-30

--- .class #id

## Plot

```{r mental-health, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(menthlth)) + geom_histogram()
```

--- .class #id

## Poor Physical or Mental Health

- `poorhlth` :  During the past 30 days, for about how many days did poor physical or mental health keep you from doing your usual
activities, such as self-care, work, or recreation?
- Values: 0-30

--- .class #id

## Plot

```{r poor-health, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(poorhlth)) + geom_histogram()
```



--- .class #id

## Imputed Race

- `imprace` : Imputed race/ethnicity value (This value is the reported race/ethnicity or an imputed race/ethnicity, if the respondent
refused to give a race/ethnicity. The value of the imputed race/ethnicity will be the most common race/ethnicity response for that region of the state) 
    - White, Non-Hispanic
    - Black, Non-Hispanic
    - Asian, Non-Hispanic
    - American Indian/Alaska Native, Non-Hispanic
    - Hispanic
    - Other Race, Non-Hispanic

--- .class #id

## Plot

```{r race, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(imprace, fill=imprace)) + geom_bar()
```


--- .class #id

## Insurance

- `insurance` : Do you have any kind of health care coverage, including health insurance, prepaid plans such as HMOs, or government
plans such as Medicare, or Indian Health Service?
    - Yes
    - No

--- .class #id

## Plot

```{r insurance, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(insurance, fill=insurance)) + geom_bar()
```


--- .class #id

## Gender

- `trnsgndr` :  Do you consider yourself to be transgender? (If yes, ask “Do you consider yourself to be male-to-female, female-to-male, or gender non-conforming?)
    - Yes, mtf
    - Yes, ftm
    - Yes, non-conforming
    - No

--- .class #id

## Plot

```{r gender, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(trnsgndr, fill=trnsgndr)) + geom_bar()
```


--- .class #id

## Gender Binary

- `trnsgndr_bin` :  Do you consider yourself to be transgender? (If yes, ask “Do you consider yourself to be male-to-female, female-to-male, or gender non-conforming?)
    - Yes
    - No

--- .class #id

## Plot

```{r gender-bin, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(trnsgndr_bin, fill=trnsgndr_bin)) + geom_bar()
```

--- .class #id

## Sexual Orientation

- `sxorient` :  Do you consider yourself to be: (We ask this question in order to better understand the health and health care needs of
people with different sexual orientations.)
    - Heterosexual
    - Homosexual
    - Bisexual
    - Other

--- .class #id

## Plot

```{r sex-orient, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(sxorient, fill=sxorient)) + geom_bar()
```


--- .class #id

## Sexual Orientation Binary

- `sxorient_bin` :  Do you consider yourself to be: (We ask this question in order to better understand the health and health care needs of
people with different sexual orientations.)
    - Heterosexual
    - Other

--- .class #id

## Plot

```{r ex-orient-bin, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(sxorient_bin, fill=sxorient_bin)) + geom_bar()
```

--- .class #id

## Life Satisfaction

- `lsatisfy`: In general, how satisfied are you with life?
    - Very Satisfied
    - Satisfied
    - Disatisfied
    - Very Disatisfied

--- .class #id

## Plot

```{r life-satisfcation, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(lsatisfy, fill=lsatisfy)) + geom_bar()
```



--- .class #id

## Life Satisfaction Binary

- `lsatisfy`: In general, how satisfied are you with life?
    - Satisfied
    - Disatisfied



--- .class #id

## Plot

```{r life-satisfaction-bin, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(lsatisfy_bin, fill=lsatisfy_bin)) + geom_bar()
```



--- .class #id

## Emotional Support

- `emtsuprt`: How often do you get the social and emotional support you need?
    - Always
    - Usually
    - Sometimes
    - Rarely
    - Never


--- .class #id

## Plot

```{r emtsupport, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(emtsuprt, fill=emtsuprt)) + geom_bar()
```



--- .class #id

## Emotional Support Binary

- `emtsuprt_bin`: How often do you get the social and emotional support you need?
    - Always/Usually
    - Sometimes/Rarely/Never


--- .class #id

## Plot

```{r emtsupport-bin, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(emtsuprt_bin, fill=emtsuprt_bin)) + geom_bar()
```



--- .class #id

## Medical Cost Barrier

- `medcost`: Was there a time in the past 12 months when you needed to see a doctor but could not because of cost?
    - Yes
    - No


--- .class #id

## Plot

```{r medcost, echo=FALSE}
library(ggplot2)
ggplot(brfss2, aes(medcost, fill=medcost)) + geom_bar()
```


---  .segue bg:grey

# Relationships

--- .class #id

## Unhealthy Days  vs Insurance

```{r, echo=F}
brfss2 %>%
  group_by(insurance) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r unhealthy-vs-insurance, echo=F}
ggplot(brfss2, aes(x=insurance, y=unhealthy.days)) + geom_boxplot()
```


--- .class #id

## Unhealthy Days  vs Medical Cost

```{r, echo=F}
brfss2 %>%
  group_by(medcost) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r medcost-vs-insurance, echo=F}
ggplot(brfss2, aes(x=medcost, y=unhealthy.days)) + geom_boxplot()
```



--- .class #id

## Unhealthy Days  vs Emotional Support

```{r, echo=F}
brfss2 %>%
  group_by(emtsuprt_bin) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r emtsupport-vs-insurance, echo=F}
ggplot(brfss2, aes(x=emtsuprt_bin, y=unhealthy.days)) + geom_boxplot()
```



--- .class #id

## Unhealthy Days  vs Gender

```{r, echo=F}
brfss2 %>%
  group_by(trnsgndr) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r gender-vs-insurance, echo=F}
ggplot(brfss2, aes(x=trnsgndr, y=unhealthy.days)) + geom_boxplot()
```


--- .class #id

## Unhealthy Days  vs Gender

```{r, echo=F}
brfss2 %>%
  group_by(trnsgndr_bin) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r genderbin-vs-insurance, echo=F}
ggplot(brfss2, aes(x=trnsgndr_bin, y=unhealthy.days)) + geom_boxplot()
```



--- .class #id

## Unhealthy Days  vs Race

```{r, echo=F}
brfss2 %>%
  group_by(imprace) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r race-vs-insurance, echo=F}
ggplot(brfss2, aes(x=imprace, y=unhealthy.days)) + geom_boxplot()
```


--- .class #id

## Unhealthy Days by General Health

```{r, echo=F}
brfss2 %>%
  group_by(genhlth) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r genhlth-vs-insurance, echo=F}
ggplot(brfss2, aes(x=genhlth, y=unhealthy.days)) + geom_boxplot()
```


--- .class #id

## Unhealthy Days by General Health

```{r, echo=F}
brfss2 %>%
  group_by(genhlth_bin) %>%
  summarise(mean=mean(unhealthy.days, na.rm=T), median=median(unhealthy.days, na.rm=T), sd = sd(unhealthy.days, na.rm=T))
```


--- .class #id

```{r genhlthbin-vs-insurance, echo=F}
library(gridExtra)
p1 <- ggplot(brfss2, aes(x=genhlth_bin, y=unhealthy.days)) + geom_boxplot()
p2 <- ggplot(brfss2, aes(x=genhlth, y=unhealthy.days)) + geom_boxplot()
grid.arrange(p1,p2, ncol=2, nrow=1)
```



--- .class #id

## Question: Does Reported General Health imply Number of Unhealthy Days

- Consider the data

```{r, echo=FALSE}
brfss2 %>%
    group_by(genhlth_bin) %>%
    summarise(mean(unhealthy.days, na.rm=T))

#Difference

16.2 - 3.95
```


--- .class #id

## Hypothesis Testing: What is Happening?

- Self Reported Health predicts number of days unhealthy. 
- Groups Differ at Baseline
- Random Chance

--- .class #id

## Groups Differ at Baseline

- We may have noticed that most comparisons in here were not showing a strong connection of unhealthy days. 
- This means that with the data we have, there may not be a problem with groups differing at baseline. 
- On the exam, you can agree with a statement like this or show otherwise. 
- If you were to show otherwise, show a connection between another variable and the outcome (unhealthy days) and that variable and the exposure (self reported health)

--- .class #id

## Hypothesis Testing: What is Happening?

- Self Reported Health predicts number of days unhealthy. 
- <delete>Groups Differ at Baseline</delete>
- Random Chance


--- .class #id

## Random Chance

- We will simulate this. 
- Based on Probabilities

```{r, eval=F}
brfss2 %>% 
    group_by(genhlth_bin) %>% 
    summarise(n=n()) %>% 
    mutate(freq=n/sum(n))
```

--- .class #id

## Random Chance

- We will simulate this. 
- Based on Probabilities

```{r, echo=F}
brfss2 %>% 
    group_by(genhlth_bin) %>% 
    summarise(n=n()) %>% 
    mutate(freq=n/sum(n))
```

--- .class #id

## Simplify Data

```{r}
brfss3 <- brfss2 %>%
          select(unhealthy.days, genhlth_bin)
```

--- .class #id

## Simulations

- One simulation
```{r}
library(tidyverse)
T = dim(brfss2)[1]
health.sim = replicate(T, sample(c("Excell/VG/G", "Fair/Poor"), size=1, prob=c(0.795, 0.205)))
health.sim = enframe(health.sim, name=NULL, value="health.sim")
brfss3 <- bind_cols(brfss3, health.sim)
test <- brfss3 %>% group_by(health.sim) %>% summarise(m=mean(unhealthy.days, na.rm=TRUE))
difference = test$m[1] - test$m[2]
difference
```

--- .class #id

## Simulations

```{r}
brfss3 <- brfss2 %>%
          select(unhealthy.days, genhlth_bin)
          
diff_sim <- function(data){

health.sim = replicate(T, sample(c("Excell/VG/G", "Fair/Poor"), size=1, prob=c(0.795, 0.205)))
health.sim = enframe(health.sim, name=NULL, value="health.sim")
data <- bind_cols(data, health.sim)
test <- data %>% group_by(health.sim) %>% summarise(m=mean(unhealthy.days, na.rm=TRUE))
difference = test$m[1] - test$m[2]
return(difference)
}

diff = replicate(1000, diff_sim(brfss3))
```


--- .class #id

## Histogram of Simulated Differences

```{r health-sim-hist, echo=F, fig.height=5}
ggplot(data=NULL, aes(diff)) + geom_histogram(binwidth = 0.1)
```

--- .class #id

## How Likely is our value?

```{r}
mean(diff>12.25)
```




--- .class #id

## Hypothesis Testing: What is Happening?

- Self Reported Health predicts number of days unhealthy. 
- <delete>Groups Differ at Baseline</delete>
- <delete>Random Chance</delete>
- **Note there is a possibilty that the number of days was asked first prompting their answer for the general health as well.** 









