---
title       : Introduction to Hypothesis Testing
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}
---  .segue bg:grey


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

```{r, echo=F}
require(tidyverse)
require(gapminder)
```

# Introduction to Hypothesis Testing

--- .class #id

## Introduction to Hypothesis Testing




```{r, echo=F}
library(tidyverse)
library(haven)
nhanes <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2009-2010/CBQPFA_F.XPT")
nhanes2 <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2009-2010/HSQ_F.XPT")
nhanes3 <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2009-2010/INQ_F.XPT")
nhanes4 <- read_xpt("https://wwwn.cdc.gov/Nchs/Nhanes/2009-2010/SMQ_F.XPT")
data <- full_join(nhanes, nhanes2, by="SEQN")
data <- full_join(data, nhanes3, by="SEQN")
data <- full_join(data, nhanes4, by="SEQN")

data <- data %>%
    select(SEQN, IND235, CBD740, SMQ020, HSD010 , CBD630) %>%
    filter(IND235 < 77 & CBD740<3 & SMQ020 <7 & HSD010 <7, CBD630<16) %>%
    mutate(health = as.factor(HSD010)) %>%
    mutate(organic = as.factor(CBD740)) %>%
    mutate(income = as.factor(IND235)) %>%
    mutate(smoke = as.factor(SMQ020))  %>%
    mutate(veggies = as.factor(CBD630))

organic <- data %>%
    mutate(health = fct_recode( health, 
        Poor =  "5", 
        Fair = "4",
        Good = "3",
        `Very Good` = "2",
        Excellent = "1"
        )) %>%
        mutate(health.bin = fct_collapse( health, 
        `Poor/Fair/Good` =  c("Poor", "Fair", "Good"),
        `Very Good/Excellent` = c("Very Good", "Excellent")
        )) %>%
    mutate(smoke = fct_recode( smoke, 
        Yes = "1", 
        No = "2"
        )) %>%   
    mutate(organic = fct_recode( organic, 
        Yes = "1", 
        No = "2"
        )) %>%
    mutate(income = fct_recode( income, 
        `199.5` = "1", 
        `599.5` = "2", 
        `1024.5` = "3",
        `1449.5` = "4", 
        `1874.5` = "5", 
        `2499.5` = "6", 
        `3324.5` = "7", 
        `4174.5` = "8", 
        `4999.5` = "9", 
        `5824.5` = "10", 
        `7324.5` = "11", 
        `9000` = "12"
        )) %>%
    mutate(income = as.numeric(as.character(income))) %>%
    mutate(veggies = as.numeric(as.character(veggies)))


```

- In most statistics courses you get to hypothesis testing at the end of the course. 
- We will do things a little different. 
- We should consider important concepts when trying to understand testing hypothesis. 

--- .class #id

## What is hypothesis Testing? 

- Many times we want to be able to compare things. 
- We wish to test whether something is better than something else. 
    - Should you buy computer "A" vs computer "B". 
- We also wish to test the validity of statements. 
    - A friend says they can force a coin to get Heads.
    - You give them a coin and they flip three times in a row and get heads. 
    - Do you believe them?

--- .class #id

## Coin Toss

```{r}

coin <- c(1, 0)

flips <- replicate(100000, sum(sample(coin, 3, replace=T)))
tab <- table(flips)
prop.table(tab)
```


--- .class #id

## Testing Our Friend

- The probability of 3 heads in 3 coin flips is 12.5%
- We may not believe that they can force heads. 
- Many times our hypothesis testing is about different groups. 


--- .class #id

## Big Question

- Does eating organic food improve your health? 
- Consider some evidence

--- .class #id

## Data Study 1

- This data comes from [NHANES: National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/index.htm)
- Large national random sample
- 2009 – 2010 data
- n = 3777

--- .class #id

## Organic Food

"In the past 30 days, did you buy any food that had the word 'organic' on the package?"

```{r organic, echo=FALSE, fig.height=5}
ggplot(organic, aes(x=organic, fill=organic)) + 
      geom_bar() + 
      xlab("Bought Organic?") + 
      theme_minimal() + 
      theme(legend.position = "none")

```


--- .class #id

## Current Health Status

"Would you say your health in general is Excellent, Very Good, Good, Fair or Poor"

```{r health-all, echo=FALSE, fig.height=5}
ggplot(organic, aes(x=health, fill=health)) + 
      geom_bar() + 
      xlab("Current Health Status")+ 
      theme_minimal() + 
      theme(legend.position = "none")

```


--- .class #id

## Current Health Status

```{r health, echo=FALSE, fig.height=5}
ggplot(organic, aes(x=health.bin, fill=health.bin)) + 
      geom_bar() + 
      xlab("Current Health Status")+ 
      theme_minimal() + 
      theme(legend.position = "none")

```


--- .class #id

## Health by Organic

```{r health-organic, echo=F, fig.height=5}
ggplot(organic, aes(x=organic, fill=health.bin)) + 
      geom_bar(position="dodge") + 
      xlab("Bought Organic?")+ 
      labs(fill="Health") + 
      theme_minimal() 
      
```



--- .class #id

## The Difference

- We could consider the difference between healthy with organic food vs healthy with conventional food. 

```{r, echo=F}
organic %>% group_by(health.bin, organic) %>% summarise(n=n()) %>% mutate(freq = n/sum(n))

0.452 - 0.325
```

--- .class #id

## Evaluating Evidence

- In this data, people who bout organic are healthier. 
- Possible Explanations
    - Eating organic improves health.
    - Groups Differed at Baseline.
    - <del>Random Chance.</del> 

--- .class #id

## Evaluating Evidence

- In this data, people who bout organic are healthier. 
- Possible Explanations
    - Eating organic improves health.
    - Groups Differed at Baseline.
    - <del>Random Chance.</del> 
- P-value: <0.00000000000002 (Will Explain Later)

--- .class #id

## People Who Buy Organic Make More Money

```{r, echo=F}
ggplot(organic, aes(y=income, x= organic, fill=organic)) + geom_boxplot() + theme_minimal() + xlab("Bought Organic?") + ylab("Monthly Income ($)") + 
      theme(legend.position = "none")
```



--- .class #id


## People Make More are Healthier

```{r rich-box, echo=F}
ggplot(organic, aes(y=income, x= health, fill=health)) + geom_boxplot() + theme_minimal() + xlab("Bought Organic?") + ylab("Monthly Income ($)") + 
      theme(legend.position = "none")
```

--- .class #id


## People Make More are Healthier



```{r rich-line, echo=F, fig.height=5}
new_plot <- organic %>% 
    group_by(income, health.bin) %>%
    summarise(n= n()) %>%
    mutate(freq=n/sum(n)) %>%
    filter(health.bin=="Very Good/Excellent") 
new_plot2 <- organic %>% group_by(income) %>% summarise(n2=n())
full_join(new_plot, new_plot2, by="income") %>%
    ggplot(aes(x=income, y=freq)) + geom_point(aes(size=n2)) + geom_line() + ylab("P(Very Good/Excellent Health") + xlab("Monthly Income ($)") + theme_minimal()


    
   
```



--- .class #id

## People Make More are Healthier


```{r top-bottom, echo=F, fig.height=5}

library(gridExtra)

top <- ggplot(organic, aes(y=income, x= organic)) + geom_boxplot() + coord_flip() + theme_minimal()
bottom <- full_join(new_plot, new_plot2, by="income") %>%
    ggplot(aes(x=income, y=freq)) + geom_point(aes(size=n2)) + geom_line() + ylab("P(Very Good/Excellent Health") + xlab("Monthly Income ($)") + theme_minimal()

grid.arrange(top, bottom, nrow=2)
```


--- .class #id


## Evaluating Evidence

- In this data, people who bout organic are healthier. 
- Possible Explanations
    - Eating organic improves health.
    - Groups Differed at Baseline.
    - <del>Random Chance.</del> 
- **We cannot make a decision here with their being more than one option**

--- .class # id

## Causal Claims

- We cannot make causal claims with groups that are not comparable at baseline. 
- We can look within similar groups to compare though. 


--- .class #id

## Line Plots by Organic vs Conventional

```{r line-organ-conv, echo=F, fig.height=5}

new_plot <- organic %>% 
    group_by(income,organic,  health.bin) %>%
    summarise(n= n()) %>%
    mutate(freq=n/sum(n)) %>%
    filter(health.bin=="Very Good/Excellent") 
new_plot2 <- organic %>% group_by(income, health.bin) %>% summarise(n2=n())
full_join(new_plot, new_plot2, by="income") %>%
    ggplot( aes(x=income, y=freq, color=organic)) + geom_point(aes(size=n2)) + geom_line() + ylab("P(Very Good/Excellent Health") + xlab("Monthly Income ($)") + theme_minimal() 

```


--- .class #id

## Evidence Enough?

- Does this satisfy our thoughts? 




--- .class #id

## Health by Smoking

- Poople who buy organic are less likely to smoke

```{r health-smoke, echo=F, fig.height=4.5}
ggplot(organic, aes(x=organic, fill=smoke)) + 
      geom_bar(position="dodge") + 
      xlab("Current Health Status")+ 
      theme_minimal() 
```

--- .class #id

## Health by Veggies

```{r health-veggies, echo=FALSE, fig.height=5}
ggplot(organic, aes(x=organic, y=veggies, color=health.bin)) + 
      geom_boxplot() + 
      xlab("Current Health Status")+ 
      theme_minimal() 
```

--- .class #id

## Conclusions

- Cannot make comparisons on non-comparable groups. 
- What can happen? 
     - Baseline difference shifts effect. 
     - Baseline difference reverses effect. 
     - Baseline difference masks true effect. 
     - Baseline difference creates false effect. 


--- .class #id

## Still Not Convinced?

- Let's simulate organic based on income only. 
- Then organc food choice is only associated with income and not with health directly. 

--- .class #id

## Simulation

```{r, echo=F}

set.seed(12345)

organic_pct <- organic %>%
        group_by(income, organic) %>%
        summarize(n=n()) %>%
        mutate(pct = n/sum(n)) %>%
        select(pct) %>%
        unlist()
        
organic_pct <- organic_pct[25:48]

organic$organic.sim = NA
for (i in 1:nrow(organic)) {
    if (organic$income[i]==199.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[1], organic_pct[2]) )  
    } else if (organic$income[i]==599.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[3], organic_pct[4]) )  
    } else if (organic$income[i]==1024.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[5], organic_pct[6]) )  
    } else if (organic$income[i]==1449.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[7], organic_pct[8]) )  
    } else if (organic$income[i]==1874.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[9], organic_pct[10]) )  
    } else if (organic$income[i]==2499.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[11], organic_pct[12]) )  
    } else if (organic$income[i]==3324.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[13], organic_pct[14]) )  
    } else if (organic$income[i]==4174.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[15], organic_pct[16]) )  
    } else if (organic$income[i]==4999.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[17], organic_pct[18]) )  
    } else if (organic$income[i]==5824.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[19], organic_pct[20]) )  
    } else if (organic$income[i]==7324.5) {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[21], organic_pct[22]) )  
    } else  {
    organic$organic.sim[i] = sample(c("Yes", "No"), 1, prob=c(organic_pct[24], organic_pct[24]) )  
    } 
}   


```



--- .class #id 


## Health by Organic Sim

- Difference exists even though we made up organic food purchases. 

```{r health-organic-sim, echo=F, fig.height=5, fig.width=13}

organic <- organic %>%
        mutate(organic.sim = as.factor(organic.sim)) %>%
        mutate(organic.sim = factor(organic.sim, levels= c("Yes", "No")))
p1 <- ggplot(organic, aes(x=organic, fill=health.bin)) + 
      geom_bar(position="dodge") + 
      xlab("Bought Organic?")+ 
      labs(fill="Health") + 
      theme_minimal() + 
      ggtitle("Original Data")
p2<- ggplot(organic, aes(x=organic.sim, fill=health.bin)) + 
      geom_bar(position="dodge") + 
      xlab("Bought Organic?")+ 
      labs(fill="Health") + 
      theme_minimal() + 
      ggtitle("Simulated Data")

grid.arrange(p1, p2, nrow=1)
```


--- .class #id

## Simulated Line Plots by Organic vs Conventional

```{r line-organsim-conv, echo=F, fig.height=5, fig.width=13}

new_plot <- organic %>% 
    group_by(income,organic,  health.bin) %>%
    summarise(n= n()) %>%
    mutate(freq=n/sum(n)) %>%
    filter(health.bin=="Very Good/Excellent") 
new_plot2 <- organic %>% group_by(income, health.bin) %>% summarise(n2=n())
p1 <- full_join(new_plot, new_plot2, by="income") %>%
    ggplot( aes(x=income, y=freq, color=organic)) + geom_point(aes(size=n2)) + geom_line() + ylab("P(Very Good/Excellent Health") + xlab("Monthly Income ($)") + theme_minimal() + ggtitle("Original Data")

new_plot <- organic %>% 
    group_by(income,organic.sim,  health.bin) %>%
    summarise(n= n()) %>%
    mutate(freq=n/sum(n)) %>%
    filter(health.bin=="Very Good/Excellent") 
new_plot2 <- organic %>% group_by(income, health.bin) %>% summarise(n2=n())
p2 <- full_join(new_plot, new_plot2, by="income") %>%
    ggplot( aes(x=income, y=freq, color=organic.sim)) + geom_point(aes(size=n2)) + geom_line() + ylab("P(Very Good/Excellent Health") + xlab("Monthly Income ($)") + theme_minimal() + ggtitle("Simulated Data")

grid.arrange(p1, p2, nrow=1)
```


--- .class #id

## What do we do? 

- Ideally We randomize at baseline so that we can make causal claims. 
- If not, there are methods to determine causal inference but they are much more difficult. 


--- .class #id

## Data Problem #2: Fruit Flies

- Fruit flies **randomly** divided into two groups of 500 each
- One group fed organic food, the other conventional food
- Measured longevity, fertility, stress resistance, activity
-  Study done by a 16-year-old girl for a science project!
- [Organically Grown Food Provides Health Benefits to Drosophila melanogaster](http://journals.plos.org/plosone/article?id=info:doi/10.1371/journal.pone.0052988)


--- .class #id

## Longevity by Organic Food

- *Data simulated to be similar to paper*

```{r fruit-flies, echo=F, fig.height=5}
set.seed(1234)
organic <- rlnorm(250, 2.95,0.26)
conventional <- rnorm(250, 15.5, 5.4)
data2 <- as_tibble(cbind(organic, conventional))
data3 <- gather(data2, key="Food", value="days")

ggplot(data=data3, aes(y=days, fill=Food)) + geom_boxplot()
```



--- .class #id

## Evaluating Evidence

- Fruit flies who eat organic food live longer in this sample. 
- Mean: $\bar{Y}_O - \bar{Y}_C = 4.23$
- Possible reasons?
    - Eating organic increases longevity. 
    - <delete>The groups differed at baseline</delete>
    - Just random chance. 


--- .class #id

## How do we check random chance?

- We can simulate this. 
- The next slide will detail the steps of this. 

--- .class #id

## Simulations

1. Assume no difference in days survived, regardless of food. 
2. Mimic random chance: Randomize into groups.
3. Compute the difference in means. 
4. Repeat the steps many times. 

--- .class #id

## Simulations

Start with the data:

```{r}
data3
```

--- .class #id

## What do we see?

- 500 rows of data. 
- Food in one column. 
- Days survived in another. 
- We can simulated 500 organic or conventional values. 

--- .class #id

## Simulations

- One simulation
```{r}
T = 500
Food.sim = replicate(T, sample(c("organic", "conventional"), 1))
Food.sim = enframe(Food.sim, name=NULL, value="Food.sim")
data3 <- bind_cols(data3, Food.sim)
test <- data3 %>% group_by(Food.sim) %>% summarise(m=mean(days))
difference = test$m[1] - test$m[2]
difference
```

--- .class #id

## Simulations

```{r}

diff_sim <- function(data){
new.col = replicate(500, sample(c("organic", "conventional"), 1))
new.col2 = enframe(new.col, name=NULL, value="Food.sim")
data <- bind_cols(data, new.col2)
test <- data %>% group_by(Food.sim) %>% summarise(m=mean(days))
difference = test$m[1] - test$m[2]
return(difference)
}

diff = replicate(1000, diff_sim(data3))
```


--- .class #id

## Histogram of Simulated Differences

```{r fruit-hist, echo=F, fig.height=5}
ggplot(data=NULL, aes(diff)) + geom_histogram(binwidth = .05)
```

--- .class #id

## How Likely is our value?

```{r}
mean(diff>4.231288)
```


--- .class #id

## Evaluating Evidence

- Possible reasons?
    - Eating organic increases longevity. 
    - <delete>The groups differed at baseline</delete>
    - <delete>Just random chance.</delete>

--- .class #id

## Conclusion

- Eating organic lenthens life. 
- **For a fruit fly**

--- .class #id

## Hypothesis Testing

- With testing we are interested in comparing different groups and having a mathematical way to judge differences. 
- Traditionally this is done with 2 hypothesis:
    - **Null Hypothesis**
    - **Alternative Hypothesis**

--- .class #id

## Hypothesis

- **Null Hypothesis**: This is what we assume to be true. Typically:
      - Suggest no difference between groups. 
      - Set population value equal to some known quantity. 
- **Alternative Hypothesis**: This is what we are testing and what we believe about the population. 
    - We must provide enough evidence that this is true. 
    

--- .class #id

## Hypothesis Testing

- It is important to understand that we are testing an unknown value in the ***population*** using ***sample*** data. 
- This means we need to have a way to approximate the population distribution from the sampling distribution. 
- We will explore sampling distributions in the next set of notes. 


--- .class #id

## Consider US Court System

- In the US a person is presumed innocent until proven guilty. 
- This is also apart of the Unitd Nations Declaration of Human Rights. 
- The goal of the prosecutor is to present enough evidence that they jury believes guilt beyond a reasonable doubt. 
- The goal of the defense is to display the doubt. 

--- .class #id

## Hypothesis Testing as Court Case

- In science we also do something similar. 
- As a scientist many times you take the role of prosecution and your goal is to make a claim and provide enough evidence so people will believe your claim. 
- Other times as reviewers and commentors we play defense and suggest flaws in the evidence and reasons not to believe the research. 




